---
title: "01-packages"

format:
  html: default
  revealjs:
    output-file: 01-packages-reveal.html
---

# R Packages {background-color="#e4282e"}

## Agenda

1.  R Packages
2.  Installation
3.  Loading

## R packages

-   **Packages** are the fundamental units of reproducible R code. They include reusable R functions, the documentation that describes how to use them, and sample data [^1]
-   People use R for lots of different tasks (data management, statistical analysis, visualization, genomics, etc.). Packages help make these different things easier/possible.
-   As of Sept 2023, there are \~20,700 R packages available on **CRAN** (the Comprehensive R Archive Network)[^2]

[^1]: Wickham and Bryan, [R Packages](https://r-pkgs.org/)

[^2]: [CRAN contributed packages](https://cran.r-project.org/web/packages/)

## Package Installation

- Packages are installed with the `install.packages` function (do this once per package)
- Loaded with the `library` function (once per session)

```{r eval=FALSE}
install.packages("package_name")
library(package_name)
```

## Install `tidyverse`

For this training, we'll primarily use one suite of packages. We'll discuss more about what this includes later, but for now, let's just install and load `tidyverse`.

. . .

This only has to be run a single time, so it's best to do it directly in your console:

```{r, eval=FALSE}
install.packages("tidyverse")
```

. . .

But, this has to be done every time you're working with code that requires the tidyverse, so best to add this to that RScript file you created.

```{r}
library(tidyverse)
```

^ Be sure to still run the code above in your console after typing it in your script.
