---
title: "10-tables"

format:
  html: default
  revealjs:
    output-file: 10-tables-reveal.html
---

# Tables {background-color="#e4282e"}

## Agenda

- `gt`
- basic table examples

## Setup 

```{r}
library(tidyverse)
library(gt)
library(palmerpenguins)
```

## Parts of a Table

![](images/gt_parts_of_a_table.svg)
Image Source: [gt.rstudio](https://gt.rstudio.com/)

## Workflow

![](images/gt_workflow_diagram.svg)
Image Source: [gt.rstudio](https://gt.rstudio.com/)

## Table Example

```{r}
# Define the start and end dates for the data range
start_date <- "2010-06-07"
end_date <- "2010-06-14"

# Create a gt table based on pre-processed
# `sp500` table data
sp500 |>
  dplyr::filter(date >= start_date & date <= end_date) |>
  dplyr::select(-adj_close) |>
  gt() |>
  tab_header(
    title = "S&P 500",
    subtitle = glue::glue("{start_date} to {end_date}")
  ) |>
  fmt_currency() |>
  fmt_date(columns = date, date_style = "wd_m_day_year") |>
  fmt_number(columns = volume, suffixing = TRUE)
```

Source: [gt.rstudio](https://gt.rstudio.com/)

## Penguins table

```{r}
# get data into format for table
pen_table <- penguins |> 
  select(species, sex, body_mass_g) |>
  filter(!is.na(sex)) |>
  group_by(species, sex) |>
  summarise(ci = list(mean_cl_normal(body_mass_g) |>
                        rename(mean=y, lwr=ymin, upr=ymax))) |> 
  unnest(cols = c(ci)) |>
  mutate(`Mean (95% CI)` = paste0(round(mean, 0), " (", round(lwr, 0), ", ",  round(upr, 0), ")")) |>
  select(-mean, -lwr, -upr) |> 
  ungroup()

pen_table
```

. . . 

```{r}
# make table
pen_table |> 
  group_by(species) |>
  gt() |> 
  tab_header(title = "Average Weight by Sex",
             subtitle = "2007-2009")
```


## Practice

Create a basic table using `dplyr` and `gt` to summarize something other than average weight using the penguins data.

::: aside
Unmute to ask or put any questions in the chat. When completed, respond to the zoom poll.
:::
