---
title: "11-data"

format:
  html: default
  revealjs:
    output-file: 11-data-reveal.html
---

# 

## Agenda

## Setup

```{r}
library(tidyverse)
library(here) 
library(skimr) 
```


## A Brief Aside on filepaths

- relative paths | from your current working directory (.RProj)
- absolute paths | from the root directory on your computer

## The `here` package

Generally, best practice to avoid using `setwd()` and use `here()` instead.

- paths with `here()` will continue to work even if the working directory is changed
- with `here()` paths will work when anyone else on any other computer runs this code...so long as they have access to the project files

## The Data

Steps: 

1. Create a folder `data/` where you're keeping your notes for this training
2. Add the file shared in the chat to that folder
3. Read the data into R using the following

```{r}
df <- read_csv(here("data/Ascent_results_sample_pivot_v2.csv"))
```

## Take a look at the data

- `skim()`
- `View()`
- `glimpse()`/`str()`
- etc.

## Practice

Read the data in and add something you learned about these data while taking a look at the data

::: aside
Unmute to ask or put any questions in the chat. When completed, respond to the zoom poll.
:::