---
title: "12-eda"

format:
  html: default
  revealjs:
    output-file: 12-eda-reveal.html
---

## Agenda

- Just get started:
  - data operations
  - calculations
- open-ended!

## Setup

```{r}
library(tidyverse)
library(patchwork)
library(here)

df_back <- read_csv(here("data/Ascent_results_sample_pivot_v2.csv"))
```

## Data Operations

- Filter to only include samples with Fentanyl levels > 0 
- Select only columns of interest: sample, Morphine, Oxycodone, Fentanyl

```{r}
filtered_data <- df_back |>
  filter(Fentanyl > 0) |>
  select(sample, Morphine, Oxycodone, Fentanyl)
```

## Calculate ratios

```{r}
filtered_data <- filtered_data |> 
  mutate(Morphine_Oxycodone = Morphine / Oxycodone)
```

## Plot new variable
```{r}
ggplot(filtered_data, aes(x = Morphine_Oxycodone)) +
  geom_histogram()
```


## Aside: functions

- user-defined functions : for tasks that you'll carry out more than twice
- avoid copy + pasting very similar code

```{r}
# Function to calculate log10
log10_safe <- function(x) {
  ifelse(x > 0, log10(x), NA)
}
```

## Use function

```{r}
filtered_data <- filtered_data |> 
  mutate(log10_Morphine_Oxycodone = log10_safe(Morphine_Oxycodone))

```

```{r}
summarized_data <- filtered_data |>
  summarise(mean_morphine = mean(Morphine, na.rm=TRUE),
            mean_oxycodone = mean(Oxycodone, na.rm=TRUE), 
            mean_fentanyl = mean(Fentanyl, na.rm=TRUE),
            sd_morphine = sd(Morphine, na.rm=TRUE),
            sd_oxycodone = sd(Oxycodone, na.rm=TRUE), 
            sd_fentanyl = sd(Fentanyl, na.rm=TRUE))

```

## Find outliers
```{r}
high_morphine <- filtered_data |>
  filter(Morphine > (summarized_data$mean_morphine + 2 * summarized_data$sd_morphine))

high_morphine
```

## Plot relationships

```{r}
plot_all <- ggplot(filtered_data, aes(x=Morphine, y=Fentanyl)) + 
  geom_point() +
  labs(title="All Data")
plot_sub <- ggplot(high_morphine, aes(x=Morphine, y=Fentanyl)) + 
  geom_point() +
  labs(title="High Morphine")

# combine plots
plot_all + plot_sub
```

## Practice

Learn something new about this dataset that we haven't yet discussed!

::: aside
Unmute to ask or put any questions in the chat. When completed, respond to the zoom poll.
:::